Analysing Constraint Grammars with a SAT-solver

### 2016-02-26 paper/lrec/eval_futurework.tex
remove DetPos(Not)Zijn example from conflicts

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
              & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings   & 336             & 2191     & 2012    \\ \hline
\# conflicts  & 8 (1 + 5 + 2)   & 1        & 51    \\ \hline
\clock{} all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
\clock{} last rule       & 4s              & 7s          & 3min 40s    \\ \hline
% This is with just 2000ish variables for Finnish, below is with 5700

% \clock{} last rule       & 4s              & 7s          & 30min 25s    \\ \hline
% \clock{} longest rule    & ?s              & ?s                & ?min     \\ \hline
%\# variables: last rule & 33915         & 247153            &    \\ \hline

\end{tabular}
\caption{}
\label{table:res}
\end{table}

@@

\paragraph{Dutch} The Dutch grammar had two kinds of errors: rule-internal due to nonexistent tags, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

There were two genuine conflicts with rule interaction. The first one is shown below:

@@

The first one is more straight-forward: two rules, which both remove an adverb, were in an order where the first has more broad condition: remove adverb if followed by a noun. In contrast, the second rule has a stricter requirement: only remove adverb if it is preceded by a determiner; the adverb itself is ambiguous with an adjective, and followed by a noun. The proble is that the first rule removes the adverb in all possible cases, and the second will not have any chance to act. If the rules were in the opposite order, then there would be no problem.

The second conflict, shown below, is more subtle.

\begin{itemize}
\item[] 
\begin{verbatim}
SELECT DetPosNotZijn IF (1 Noun) ;
SELECT DetPos IF (-1 (vbser pres p3 sg)) 
                 (0 "zijn") (1 Noun);
\end{verbatim}
\end{itemize}

The condition of the second rule cannot be true: there can never be a ``zijn'' in the target, because the previous rule has selected everything that is a possessive determiner and not ``zijn''. 
%Not firing is not an option: if conditions of the first rule do not hold, then they will not hold for the second either. If the target is the only left, then 
% "target" in case of SELECT means complement of the mentioned target. so 
%    some of [adj, det pos zijn, verb zijn, punct, ...]  is true
%    all of  [det pos f sg, det pos m pl, det pos foo bar] are false.
% ie. there has been a rule that has removed detPosNotZijn, or they haven't been there in the first place.
%

Another remark on this formulation of a rule. 
As per VISL CG-3, the condition \texttt{(0 "zijn")} does not require that the input would have determiner and ``zijn'' in the same reading: just that there is any determiner and any ``zijn'' in the cohort. 
The first example of a rule conflict had a similar requirement, but used correctly: the intended semantics was indeed that the adjective and the adverb are in different readings. However, in this case, the correct way to express the target is \texttt{SELECT DetPos + "zijn"}.

Can we catch this additional problem with our tool? The example word constructed by the SAT-solver may as likely be one of the following:

@@

\item[b.] \begin{verbatim}
"w2"
    w2<det><pos><mfn><pl><"zijn">
\end{verbatim}
\end{itemize}

In the latter case, the grammar writer would hopefully notice the imprecise definition and change their definition. However, this is more of a happy side effect than intended feature---the grammar writer cannot count on our tool to notice all similar cases.
We could add a feature that alerts in all cases where a condition with 0 is used, or even automatically construct a version of the rule that moves whatever tags in the 0-condition into the target.

>>>>

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}

\hline
              & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings   & 336             & 2191     & 2012    \\ \hline
\# conflicts  & 7 (1 + 5 + 1)   & 1        & 51    \\ \hline
\clock{} all rules       & ??s              & ??min ??s    & ?h ??min    \\ \hline
\clock{} last rule       & ??s              & ??s          & ?min ??s    \\ \hline

%\# variables: last rule & 33915         & 247153            &    \\ \hline

\end{tabular}
\caption{\todo{make more tables}}
\label{table:res}
\end{table}

@@

\paragraph{Dutch} The Dutch grammar had two kinds of errors: rule-internal, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

There was one genuine conflict with rule interaction, shown below:

@@

Two rules, which both remove an adverb, were in an order where the first has more broad condition: remove adverb if followed by a noun. In contrast, the second rule has a stricter requirement: only remove adverb if it is preceded by a determiner; the adverb itself is ambiguous with an adjective, and followed by a noun. The proble is that the first rule removes the adverb in all possible cases, and the second will not have any chance to act. If the rules were in the opposite order, then there would be no problem.

%%% Nevermind, this wasn't a conflict, my program was just buggy :-P
%This is not a conflict, just pointing out an example how the SAT-solver constructs the only possible sentence where this works.

%\begin{itemize}
%\item[] 
%\begin{verbatim}
%SELECT DetPosNotZijn IF (1 Noun) ;
%SELECT DetPos IF (-1 (vbser pres p3 sg)) 
%                 (0 "zijn") (1 Noun);
%\end{verbatim}
%\end{itemize}

% SELECT DetPosNotZijn IF (1 Noun);
%If the first rule fires, then the condition of the second rule cannot be true: there can never be a ``zijn'' in the target, because the previous rule has selected everything that is a possessive determiner and not ``zijn''. 
%So, the first rule must not fire. It cannot be because condition does not hold: the condition (1 Noun) is shared with the second rule, and

We also tested rules individually, in a way that a grammar writer might use our tool when writing new rules.
The following rule was one of them:

\begin{itemize}
\item[] 
\texttt{SELECT DetPos IF (-1 (vbser pres p3 sg)) (0 "zijn") (1 Noun);}
\end{itemize}

As per VISL CG-3, the condition \texttt{(0 "zijn")} does not require that the input would have determiner and ``zijn'' in the same reading: just that there is any determiner and any ``zijn'' in the cohort.

\todo{Find a better example of the imprecise 0 condition --- the ambiguity class thing would eliminate this one!}

Can we catch this imprecise formulation with our tool? The example word constructed by the SAT-solver may as likely be one of the following:

@@

\item[b.] \begin{verbatim}
"w2"
    w2<det><pos><mfn><pl><"zijn">
    w2<anything>
\end{verbatim}
\end{itemize}

Incidentally, the first example of a rule conflict had a similar requirement, but used correctly: the intended semantics was indeed that the adjective and the adverb are in different readings. However, in this case, the correct way to express the target is \texttt{SELECT DetPos + "zijn"}.

In the latter case, the grammar writer would hopefully notice the imprecise definition and change their definition. However, this is more of a happy side effect than intended feature---the grammar writer cannot count on our tool to notice all similar cases.
This kind of confusion is found in other rules as well. 
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.
